### 2020-05-11 serial.tex
some thoughts and suggestions about Results

We illustrate the theory by applying it to a specific example of an epidemic model. 
We model disease spread with a renewal-equation model \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, roberts2007model, champredon2018equivalence}.
Ignoring births and deaths, changes in the proportion of susceptible individuals $S(t)$ and incidence of infection $i(t)$ can be written as:
\begin{equation}
\begin{aligned}
\frac{\mathrm{d}S}{\mathrm{d}t} &= - i(t)\\
i(t) &= \Ro S(t) \int_0^\infty i(t-\tau) \gdist(\tau) \dtau,
\end{aligned}
\label{eq:renewal}
\end{equation}
where \Ro is the basic reproduction number, and $\gdist(\tau)$ is the intrinsic generation-interval distribution (i.e., the forward generation-interval distribution of a primary case in a fully susceptible population.
Then, the forward generation-interval for a cohort of individuals that were infected at time $\psymp$ follows \citep{champredon2015intrinsic}:
\begin{equation}
\gdist_\psymp (\tau) \propto \gdist(\tau) S(\psymp + \tau),
\end{equation}
which allows us to separate the joint probability distribution $h_\psymp$ of the forward incubation period and the forward generation-interval distribution as a product of the proportion of susceptible individuals $S$ and the joint probability distribution $h$ of the forward incubation period and the intrinsic generation intervals:
\begin{equation}
h_\psymp (x, \tau) \propto h(x, \tau) S(\psymp + \tau).
\end{equation}
We further assume that the forward incubation period distributions does not vary across cohorts over the course of an epidemic, as they represent the life history of a disease, and use $k$ without subscript instead. 
Then, we have:
\begin{equation}
\begin{aligned}
k(x) &= \int_0^\infty h(x, \tau) \dtau\\
\gdist(\tau) &= \int_0^\infty h(x, \tau) \dx
\end{aligned}
\end{equation}
Finally, the case reproduction for this model is defined as follows:
\begin{equation}
\Rc(t) = \Ro \int_0^\infty \gdist(\tau) S(t+\tau) \dtau.
\end{equation}

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=0.95\textwidth]{forward.pdf}
\caption{
\textbf{Epidemiological dynamics and changes in mean forward and backward delay distributions.}
(A) Cumulative incidence over time. \jd{Would it be easier for people to think about dynamics if we show incidence instead?}
(B--D) Changes in the mean forward incubation period, generation interval, and serial interval.
(E--G) Changes in the mean backward incubation period, generation interval, and serial interval.
Black lines represent the results of a deterministic simulation.
Orange lines and points represent the average of 10 stochastic simulations;
cumulative incidence from the stochastic simulations in panel A are shown without averaging.
Forward incubation periods and intrinsic generation-intervals are assumed to be independent of each other.
See Table 1 for parameter values.
}
\label{fig:epi}
\end{center}
\end{figure}

@@

The forward serial interval distributions depend on three intervals (\fref{diagram}): (i) the backward incubation period distributions, (ii) the forward generation-interval distributions, and (iii) the forward incubation period distributions.
Since both forward incubation period (\fref{epi}B) and generation-interval (\fref{epi}C) distributions remain roughly constant, changes in the forward serial-interval distributions are predominantly driven by changes in the backward incubation period distributions, whose mean increases over time (\fref{epi}D). \jd{Are you saying that this is true in general, or for these parameters? Have we said enough about these parameters to make this clear to others?}
In this case, the increase in the mean backward incubation period directly translates to the decrease in the mean forward serial interval.
Therefore, the forward serial intervals can decrease over time even in the absence of significant depletion of the susceptible pool;
decrease in incidence via intervention strategies can still cause the backward incubation period to increase, which in turn can decrease serial intervals. 
In general, relative contributions of the three distributions depend on their shapes, correlations between incubation periods and generation intervals, and overall epidemiological dynamics.

Qualitative patterns in the mean backward delays is robust across all delay distributions because they are predominantly driven by the changes in incidence, which in turn affects cohort sizes (\fref{epi}D--F; \eref{backward}).
When incidence increases exponentially, individuals are more likely to have been infected more recently and therefore, we are more likely to observe shorter intervals (\eref{backexp}).
When incidence decreases, we are more likely to observe longer intervals for similar but opposite reasons.

@@

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{diagram.pdf}
\caption{
\textbf{Estimating the reproduction number from the observed serial intervals.}
(A) Schematic representation of line list data collected during an epidemic.
(B) Changes in the estimate of \Ro based on the observed serial intervals over time.
(C) Schematic representation of line list data rearranged by symptom onset date of infectors.
(B) Changes in the estimate of \Ro based on the cohort-averaged serial intervals over time.
Black dashed lines represent the mean initial forward serial interval and \Ro.
Black solid lines represent the mean intrinsic serial interval and \Rintrinsic.
Colored solid lines represent the mean estimates of \Ro across 10 stochastic simulations.
Colored ribbons represent the range of estimates of \Ro across 10 stochastic simulations.
}
\label{fig:obsrR}
\end{figure}

Now, we turn to practical issues.
In order to have an unbiased estimate on \Ro, we need to use serial-interval distributions estimated based on primary cohorts (i.e., infectors who share the same symptom onset time) at the early stage of the epidemic (i.e. initial forward serial-interval distribution).
However, when an epidemic is ongoing, the observed serial intervals are subject to right-censoring because we cannot observe a serial interval if either an infector or an infectee has not yet developed symptoms;
for example, if we were to measure serial intervals on Day 8 as in \fref{obsrR}A, we will only be able to observe the first 6 events (ID 1--6).
\fref{obsrR}B demonstrates how the effect of right-censoring in the observed serial intervals translates to the underestimation of the basic reproduction number \Ro in our stochastic simulations.
Notably, even if we can observe \emph{all} serial intervals across all transmission pairs after the epidemic has ended, we still underestimate the initial mean forward serial interval and therefore \Ro by a large amount because the observed serial-interval distribution converges to the intrinsic serial-interval distribution;
we still underestimate the intrinsic value slightly due to contraction of the forward generation-interval distribution during the susceptible depletion phase (\fref{epi}C).

We provide a simple, heuristic way of assessing potential biases in the estimate of the mean initial forward serial interval and therefore \Ro retrospectively.
Once serial intervals have been observed after the epidemic has been sufficiently progressed, we can rearrange the line list and group observed serial intervals based on the symptom onset date of infectors (\fref{obsrR}C).
Then, we can compare how estimates of the mean serial interval as well as \Ro change as we incorporate more recent cohorts into the analysis;
that is, we analyze observed serial intervals from infectors who became symptomatic before time $t$ and evaluate how the estimates change as we increase $t$
(\fref{obsrR}D).
During the exponential growth phase, the estimates of the mean serial interval and \Ro are consistent with the target value;
adding more data allows us to make more precise inference during this period.
However, the cohort-averaged estimates decrease rapidly soon after the exponential growth period, reflecting changes in the forward serial-interval distributions.
This approach allows us to detect dynamical changes in the forward serial-interval distributions and its effect on the estimates of \Ro.

@@

Finally, we re-analyze serial intervals of COVID-19 collected by \cite{du2020serial} from mainland China, outside Hubei province, based on transmission events reported between January 21--February 8, 2020;
\cite{du2020serial} estimated the mean serial of 3.96 days (95\% CI 3.53–4.39 days) and \Ro of 1.32 (95\% CI 1.16–1.48).
\fref{du}A shows that the mean forward serial interval decreases over time.
While the decrease is likely to be affected by the right-censoring (indicated by the closeness between the quantiles of the observed serial intervals and maximum \emph{observable} serial intervals), increase in the proportion of negative serial intervals are clearly indicative of the changes in the forward serial-interval distribution.
The decrease in the mean forward serial interval is likely to be cause by increase in the mean backward serial intervals as well as intervention efforts (e.g., case isolation will shorten the forward generation intervals).
\fref{du}B shows that the mean backward serial interval increases over time, reflecting decrease in incidence over time.

@@

\fref{du}C shows the cohort-averaged estimates of \Ro, which remain roughly constant until day 7 and suddenly decreases;
this sudden decrease is indicative of changes in the forward serial intervals as shown in our simulation study (\fref{obsrR}).
The estimates of \Ro based on the early forward serial intervals are also consistent with previous estimates of \Ro of the COVID-19 epidemic in China \citep{majumder2020early, park2020reconciling}.
We note that early cohort-averaged estimates of \Ro are unlikely to be affected by the right-censoring as we expect the degree of right-censoring to be low (\fref{du}A).
This example clearly demonstrates the danger of using the observed serial intervals to calculate the basic reproduction number without integrating serial intervals into cohorts..

>>>>

We illustrate the theory by applying it to a specific example of an epidemic model. 
We model disease spread with a renewal-equation model \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, roberts2007model, champredon2018equivalence}.
Ignoring births and deaths, changes in the proportion of susceptible individuals $S(t)$ and incidence of infection $i(t)$ can be written as:
\begin{equation}
\begin{aligned}
\frac{\mathrm{d}S}{\mathrm{d}t} &= - i(t)\\
i(t) &= \Ro S(t) \int_0^\infty i(t-\tau) \gdist(\tau) \dtau,
\end{aligned}
\label{eq:renewal}
\end{equation}
where \Ro is the basic reproduction number, and $\gdist(\tau)$ is the intrinsic generation-interval distribution (i.e., the forward generation-interval distribution of a primary case in a population where the number of susceptibles remains constant).
Then, the forward generation-interval for a cohort of individuals that were infected at time $\psymp$ follows \citep{champredon2015intrinsic}:
\begin{equation}
\gdist_\psymp (\tau) \propto \gdist(\tau) S(\psymp + \tau),
\end{equation}
which allows us to separate the joint probability distribution $h_\psymp$ of the forward incubation period and the forward generation-interval distribution as a product of the proportion of susceptible individuals $S$ and the joint probability distribution $h$ of the forward incubation period and the intrinsic generation intervals:
\begin{equation}
h_\psymp (x, \tau) \propto h(x, \tau) S(\psymp + \tau).
\end{equation}
We further assume that the forward incubation period distributions does not vary across cohorts over the course of an epidemic, as they represent the life history of a disease, and use $k$ without subscript instead. 
Then, we have:
\begin{equation}
\begin{aligned}
k(x) &= \int_0^\infty h(x, \tau) \dtau\\
\gdist(\tau) &= \int_0^\infty h(x, \tau) \dx
\end{aligned}
\end{equation}
Finally, the case reproduction for this model is defined as follows:
\begin{equation}
\Rc(t) = \Ro \int_0^\infty \gdist(\tau) S(t+\tau) \dtau.
\end{equation}

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=0.95\textwidth]{forward.pdf}
\caption{
\textbf{Epidemiological dynamics and changes in mean forward and backward delay distributions.}
(A) Cumulative incidence over time. \jd{Would it be easier for people to think about dynamics if we show incidence instead?}
(B--D) Changes in the mean forward incubation period, generation interval, and serial interval.
(E--G) Changes in the mean backward incubation period, generation interval, and serial interval.
Black lines represent the results of a deterministic simulation.
Orange lines and points represent the average of 10 stochastic simulations;
cumulative incidence from the stochastic simulations in panel A are shown without averaging.
\jd{I don't see orange lines, and the location of the dashed line is not explained. I guess you use the forward value at time 0? Would the intrinsic value be better?}
Forward incubation periods and intrinsic generation-intervals are assumed to be independent of each other.
See Table 1 for parameter values.
}
\label{fig:epi}
\end{center}
\end{figure}

@@

The forward serial interval distributions depend on three intervals (\fref{diagram}): (i) the backward incubation period distributions, (ii) the forward generation-interval distributions, and (iii) the forward incubation period distributions.
Since both forward incubation period (\fref{epi}B) and generation-interval (\fref{epi}C) distributions remain roughly constant, changes in the forward serial-interval distributions are predominantly driven by changes in the backward incubation period distributions, whose mean increases over time (\fref{epi}D). \jd{Are you saying that this is true in general, or for these parameters? Have we said enough about these parameters to make this clear to others?}
In this case, the increase in the mean backward incubation period directly translates to the decrease in the mean forward serial interval.
Therefore, the forward serial intervals can decrease over time even in the absence of significant depletion of the susceptible pool;
decrease in incidence via intervention strategies can still cause the backward incubation period to increase, which in turn can decrease serial intervals. 
\jd{Bringing intervention in here is potentially confusing. If you want to do it, maybe try a bit more lead in?} 
In general, relative contributions of the three distributions depend on their shapes, correlations between incubation periods and generation intervals, and overall epidemiological dynamics.

We see similar qualitative patterns in all three backward delays (\fref{epi}E--G; \eref{backward}), because they are predominantly driven by the rate of change in incidence, which in turn affects relative cohort sizes.
When incidence is increasing, individuals are more likely to have been infected recently and therefore, we are more likely to observe shorter intervals (\eref{backexp}).
Similarly, when incidence decreases, we are more likely to observe longer intervals.

@@

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{diagram.pdf}
\caption{
\textbf{Estimating the reproduction number from the observed serial intervals.}
(A) Schematic representation of line list data collected during an epidemic.
(B) Estimates of \Ro based on all observed serial intervals completed by a given time. \jd{Is this accurate?}
(C) Schematic representation of line list data rearranged by symptom onset date of infectors.
(D) Estimates of \Ro based on all observed serial intervals started by a given time. 
Black dashed lines represent the mean initial forward serial interval and \Ro.
Black solid lines represent the mean intrinsic serial interval and \Rintrinsic.
Colored solid lines represent the mean estimates of \Ro across 10 stochastic simulations.
Colored ribbons represent the range of estimates of \Ro across 10 stochastic simulations.
}
\label{fig:obsrR}
\end{figure}

Now we turn to practical issues.
In order to have an unbiased estimate of \Ro, we need to use serial-interval distributions estimated based on primary cohorts (i.e., infectors who share the same symptom onset time) at the early stage of the epidemic (i.e. initial forward serial-interval distribution).
However, when an epidemic is ongoing, the observed serial intervals are subject to right-censoring because we cannot observe a serial interval if either an infector or an infectee has not yet developed symptoms;
for example, if we were to measure serial intervals on Day 8 as in \fref{obsrR}A, we will only be able to observe the first 6 events (ID 1--6).
\fref{obsrR}B demonstrates how the effect of right-censoring in the observed serial intervals translates to the underestimation of the basic reproduction number \Ro in our stochastic simulations.
Notably, even if we could observe, and average, \emph{all} serial intervals across all transmission pairs after the epidemic has ended, we would still underestimate the initial mean forward serial interval (and therefore \Ro), likely by a large amount, because the observed serial-interval distribution converges to the intrinsic serial-interval distribution
(in fact, we would even underestimate the intrinsic value slightly due to contraction of the forward generation-interval distribution during the susceptible depletion phase (\fref{epi}C)).

Here we provide a heuristic way of assessing potential biases in the estimate of the mean initial forward serial interval and therefore \Ro retrospectively.
Once serial intervals have been observed after the epidemic has been sufficiently progressed, we can rearrange the line list and group observed serial intervals based on the symptom onset date of infectors (\fref{obsrR}C).
Then, we can compare how estimates of the mean serial interval as well as \Ro change as we incorporate more recent cohorts into the analysis;
that is, we analyze observed serial intervals from infectors who became symptomatic before time $t$ and evaluate how the estimates change as we increase $t$.
This approach is analogous to averaging over a set of forward intervals, just as using all information up to a certain time is analogous to averaging over a set of backward intervals.
(\fref{obsrR}D).
During the exponential growth phase, the estimates of the mean serial interval and \Ro are consistent with the target value;
adding more data allows us to make more precise inference during this period.
However, the cohort-averaged estimates decrease rapidly soon after the exponential growth period, reflecting changes in the forward serial-interval distributions.
This approach allows us to detect dynamical changes in the forward serial-interval distributions and its effect on the estimates of \Ro.

@@

Finally, we re-analyze serial intervals of COVID-19 collected by \cite{du2020serial} from mainland China, outside Hubei province, based on transmission events reported between January 21--February 8, 2020;
\cite{du2020serial} estimated the mean serial of 3.96 days (95\% CI 3.53–4.39 days) and \Ro of 1.32 (95\% CI 1.16–1.48).
\fref{du}A shows that the mean forward serial interval decreases over time.
While the decrease is likely to be affected by the right-censoring (indicated by the closeness between the quantiles of the observed serial intervals and maximum \emph{observable} serial intervals), increase in the proportion of negative serial intervals are clearly indicative of the changes in the forward serial-interval distribution.
\jd{Is censoring relevant to negative values as well? We might not have (or can't?) observe events before a certain time? It's a bit confusing.}
The decrease in the mean forward serial interval was likely caused by increase in the mean backward serial intervals \jd{incubation periods? is it obvious to the reader why these are increasing?} as well as intervention efforts which reduce generation intervals by preventing infections once cases are identified.
\fref{du}B shows that the mean backward serial interval increases over time, reflecting decrease in incidence over time.

@@

\fref{du}C shows the cohort-averaged estimates of \Ro, which remain roughly constant until day 7 and suddenly decreases;
this sudden decrease is indicative of changes in the forward serial intervals as shown in our simulation study (\fref{obsrR}).
The estimates of \Ro based on the early forward serial intervals are also consistent with previous estimates of \Ro of the COVID-19 epidemic in China \citep{majumder2020early, park2020reconciling}.
We note that early cohort-averaged estimates of \Ro are unlikely to be affected by the right-censoring as we expect the degree of right-censoring to be low (\fref{du}A).
This example demonstrates the danger of using the observed serial intervals to calculate the basic reproduction number without integrating serial intervals into cohorts..
